Functional Pearl: the Proof Search Monad

### 2015-10-13 paper/main.tex
More chit-chat in the paper.

\begin{figure}
  \centering
\begin{ocaml}
(* These two modules belong to the library. *)
module type LOGIC = sig     module MakeProofTree (L: LOGIC) = struct
  type formula                type derivation = L.formula * rule
  type rule_name              and rule = L.rule_name * premises
end                           and premises = Premises of derivation list
                            end

(* This is the client code using modules from the library. *)
module MyLogic = struct
  type formula = ... (* as before *)
  type rule_name = R_And | R_Refl | R_Inst
end
module MyProofTree = MakeProofTree(MyLogic)
\end{ocaml}
  \caption{The functor of proof trees (library and client code)}
  \label{fig:proof-trees}
\end{figure}

@@

\begin{figure}
  \centering
\begin{ocaml}
module WriterT (M: MONAD) (L: MONOID): sig      module L = struct
  type 'a m = (L.a * 'a) M.m                      type a = MyProofTree.derivation list
  val return: 'a -> 'a m                          let empty = []
  val ( >>= ): 'a m -> ('a -> 'b m) -> 'b m       let append = List.append
                                                end
  val tell: L.a -> unit m
end = ...

@@

\section{Generating a proof tree using the logical rules}
\label{sec:generate-logical}

>>>>

\begin{figure}
  \centering
\begin{ocaml}
(* These two modules belong to the library. *)
module type LOGIC = sig     module Derivations.Make (L: LOGIC) = struct
  type formula                type derivation = L.formula * rule
  type rule_name              and rule = L.rule_name * premises
end                           and premises = Premises of derivation list
                            end

(* This is the client code using modules from the library. *)
module MyLogic = struct
  type formula = ... (* as before *)
  type rule_name = R_And | R_Refl | R_Inst
end
module MyDerivations = Derivations.Make(MyLogic)
\end{ocaml}
  \caption{The functor of proof trees (library and client code)}
  \label{fig:proof-trees}
\end{figure}

@@

\begin{figure}
  \centering
\begin{ocaml}
module WriterT (M: MONAD) (L: MONOID): sig      module L = struct
  type 'a m = (L.a * 'a) M.m                      type a = Derivations.derivation list
  val return: 'a -> 'a m                          let empty = []
  val ( >>= ): 'a m -> ('a -> 'b m) -> 'b m       let append = List.append
                                                end
  val tell: L.a -> unit m
end = ...

@@

\section{Full example: quantifiers and derivations of logical rules}
\label{sec:generate-logical}

The final section is concerned with adding quantifiers to the language, and
generating derivation trees that reflect the rules of the \emph{logic} rather
than the rules of the \emph{algorithm}. The former is useful for debugging the
inner workings of the algorithm; the latter is useful for producing a proof
certificate.

\begin{figure}
  \centering
\begin{ocaml}
(* Update of the [MyLogic] module. *)
type formula =
  (* ... *)
  | Exists of string * formula
  | Forall of string * formula

type rule_name =
  | R_And
  | R_Refl of atom
  | R_OrL
  | R_OrR
  | R_ExistsE of atom
  | R_ForallE

(* Update of the [Derivations] module. *)
type derivation =
  goal * rule

and goal =
  L.state * L.formula

and (* ... *)

(* Update of the [Combinators] module. *)
let prove (goal: Logic.formula) (x: ('a * rule_name) m): 'a outcome =
  M.(x >>= fun (premises, (env, rule)) ->
    return (env, ((env, goal), (rule, Premises premises))))

(* Update of the client. *)
let rec solve (state: state) (goal: formula): state outcome =
  (* ... *)
  | Exists (atom, g) ->
      let var, g, state = open_flexible state atom g in
      let var = assert_open var in
      prove goal begin
        premise (solve state g) >>= fun state ->
        qed (R_ExistsE (name var state)) state
      end
\end{ocaml}
  \caption{Dealing with quantifiers}
  \label{fig:quantifiers}
\end{figure}

We augment the data type of formulas with quantifiers; we replace the type of
rules with the rules from the logic. Furthermore, we demand that the \Rule{Refl}
and \Rule{ExistsE} rules be annotated with their argument.

Bound variables are globally-unique atoms (strings); open variables are
equivalence classes of the union-find, as before.

Now, a \li+goal+ is no longer just a formula: it is a formula interpreted in a
given state. Indeed, since we now open binders and substitute bound variables
for open variables, reading a formula only makes sense in a state where the open
variables of the formula are properly bound in the environment.

We thus change the \li+prove+ combinator to record the state upon creating
a new node in the derivation tree. More precisely, the \li+prove+ combinator
records the state after the premises have been satisfied. This is exactly what
is needed to conclude a rule; for instance, in the \li+Exists+ case, the solver
prods the union-find state to discover the instantiation choice that
\emph{has been made} for the existentially-quantified variable, and records it
in the proof tree.

Here is a pretty-printed version of a derivation tree.

\begin{verbatim}
prove ∀x. ∀z. ∃y. (y = x \/ y = z) /\ y = z using [forall]
| prove ∀z. ∃y. (y = x \/ y = z) /\ y = z using [forall]
| | prove ∃y. (y = x \/ y = z) /\ y = z using [exists[z]]
| | | prove (z = x \/ z = z) /\ z = z using [/\]
| | | | prove z = x \/ z = z using [\/_r]
| | | | | prove z = z using axiom [refl[z]]
| | | | prove z = z using axiom [refl[z]]
\end{verbatim}
